Revealing the movement dynamics of hippocampal replay using state space modeling

### 2020-09-30 Article.tex

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on event significance, and thus analyze only a small subset of events (~10-45\%). As a result, we lack a clear picture of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v3}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each SWR. Each row represents a possible dynamic and each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total number of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of average tetrode multiunit spike rates for each dynamic category within each SWR.
}
\label{5}
\end{figure*}

@@

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the headstage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track (herein referred to as the track graph), where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to assign the position detected in each video frame to the most likely track segment. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is empirically estimated, and changes with each time point to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal of the state transition model to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm(cite?). After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the nearest point of the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from influencing neighboring segments inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

>>>>

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on SWR event significance, and thus analyze only a small subset of events (\texttildelow10-45\%). As a result, we lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each SWR. Each row represents a possible dynamic and each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total number of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of average tetrode multiunit spike rates for each dynamic category within each SWR.
}
\label{5}
\end{figure*}

@@

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the headstage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track (herein referred to as the track graph), where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to assign the position detected in each video frame to the most likely track segment. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is empirically estimated, and changes with each time point to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal of the state transition model to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm \cite{ViterbiErrorboundsconvolutional1967}. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the nearest point of the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from influencing neighboring segments inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.
